---
title: "Assignment 3: Data Exploration"
author: "Xuancheng Guo"
date: "Spring 2024"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Exploration.

## Directions

1.  Rename this file `<FirstLast>_A03_DataExploration.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction. 
4.  Assign a useful **name to each code chunk** and include ample **comments** with your code.
5.  Be sure to **answer the questions** in this assignment document.
6.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
7.  After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai.

**TIP**: If your code extends past the page when knit, tidy your code by manually inserting line breaks.

**TIP**: If your code fails to knit, check that no `install.packages()` or `View()` commands exist in your code. 

---

```{r}
file.rename("A03_DataExploration.Rmd", "XuanchengGuo_A03_DataExploration.Rmd")
```

## Set up your R session

1.  Check your working directory, load necessary packages (tidyverse, lubridate), and upload two datasets: the ECOTOX neonicotinoid dataset (ECOTOX_Neonicotinoids_Insects_raw.csv) and the Niwot Ridge NEON dataset for litter and woody debris (NEON_NIWO_Litter_massdata_2018-08_raw.csv). Name these datasets "Neonics" and "Litter", respectively. Be sure to include the subcommand to read strings in as factors.

```{r,  message = FALSE}
# Loading datasets
getwd()
install.packages("tydyverse")
install.packages("lubridate")
install.packages("tinytex")
library(tidyverse)
library(lubridate)
library(tinytex)
```
```{r}
# Naming datasets
Neonics <- read.csv("./Data/Raw/ECOTOX_Neonicotinoids_Insects_raw.csv", stringsAsFactors = TRUE)

Litter <- read.csv("./Data/Raw/NEON_NIWO_Litter_massdata_2018-08_raw.csv", stringsAsFactors = TRUE)
```



## Learn about your system

2.  The neonicotinoid dataset was collected from the Environmental Protection Agency's ECOTOX Knowledgebase, a database for ecotoxicology research. Neonicotinoids are a class of insecticides used widely in agriculture. The dataset that has been pulled includes all studies published on insects. Why might we be interested in the ecotoxicology of neonicotinoids on insects? Feel free to do a brief internet search if you feel you need more background information.

> Answer:Because Neonicotinoids are widely used in agriculture, we might want to understand how that influence each type of corp and how these influence might transfer to human beings. 

3.  The Niwot Ridge litter and woody debris dataset was collected from the National Ecological Observatory Network, which collectively includes 81 aquatic and terrestrial sites across 20 ecoclimatic domains. 32 of these sites sample forest litter and woody debris, and we will focus on the Niwot Ridge long-term ecological research (LTER) station in Colorado. Why might we be interested in studying litter and woody debris that falls to the ground in forests? Feel free to do a brief internet search if you feel you need more background information.

> Answer: Forest litter and woody debris are a major source of nutrients for forest system. By focusing on LTER's data, we can further understand how these nutrients influence ecosystem in Colorado.

4.  How is litter and woody debris sampled as part of the NEON network? Read the NEON_Litterfall_UserGuide.pdf document to learn more. List three pieces of salient information about the sampling methods here:

> Answer: 
 1. One pair of litter trap (one elevated and one grounded) is deployed every 400 m^2 plot area.Plot edge must be seperated 150% of one edge to another.
 2. Trap placement within plots may be either targeted or randomized, depending on the vegetation.
 3. In sites with forested tower airsheds, the litter sampling is targeted to take place in 20 40m x 40m plots. In sites with low-statured vegetation over the tower airsheds, litter sampling is targeted to take place in 4 40m x 40m tower plots (to accommodate co-located soil sampling) plus 26 20m x 20m plots.

## Obtain basic summaries of your data (Neonics)

5.  What are the dimensions of the dataset?

```{r}
dimention_Neo <- dim(Neonics)
dimention_Neo
```

6.  Using the `summary` function on the "Effect" column, determine the most common effects that are studied. Why might these effects specifically be of interest?

```{r}
summary(Neonics$Effect)
```

> Answer: The most commen effects studied is "Population", followed by "Mortality". These two are two major statistic result that can use to understand how effective is our design and treatment. 


7.  Using the `summary` function, determine the six most commonly studied species in the dataset (common name). What do these species have in common, and why might they be of interest over other insects? Feel free to do a brief internet search for more information if needed.[TIP: The `sort()` command can sort the output of the summary command...]

```{r}
common_species_studied <- summary(Neonics$Species.Common.Name)
sort(common_species_studied)
common_species_studied[1:6]
```

> Answer: The top six species studied are "Honey Bee", "Parasitic Wasp", "Buff Tailed Bumblebee", "Carniolan Honey Bee", "Bumble Bee", and "Italian Honeybee". They have multiple things in common that researchers might interested in: They all pollinate flowers and produce honey. These species also share relatively similar diat.

8.  Concentrations are always a numeric value. What is the class of `Conc.1..Author.` column in the dataset, and why is it not numeric?

```{r}
view(Neonics)
class(Neonics$Conc.1..Author.)
```

> Answer: The variable is a factor because they only represent quanlitative result rather than numeric.

## Explore your data graphically (Neonics)

9.  Using `geom_freqpoly`, generate a plot of the number of studies conducted by publication year.

```{r}
# Load tools
library(ggplot2)

ggplot(Neonics, aes(x = Publication.Year)) + geom_freqpoly()
```

10. Reproduce the same graph but now add a color aesthetic so that different Test.Location are displayed as different colors.

```{r}
ggplot(Neonics) + 
  geom_freqpoly(aes(x = Publication.Year, color = Test.Location)) 
```

Interpret this graph. What are the most common test locations, and do they differ over time?

> Answer: Our data start from 1980 to 2020, and the most common test location is Lab which increase by year until 2015ish then drop and end at 2020. Researchers also do a lot of natual field studies around 2010 while they also do all other forms of researches. 

11. Create a bar graph of Endpoint counts. What are the two most common end points, and how are they defined? Consult the ECOTOX_CodeAppendix for more information. 

[**TIP**: Add `theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))` to the end of your plot command to rotate and align the X-axis labels...]

```{r}
ggplot(Neonics) + 
  geom_bar(aes(x = Endpoint)) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

> Answer: The most common two endpoints are NOEL and LOEL. NOEL, stand for no observable effect level, and LOEL on the opposite, stand for lowest observable effect level. They stand for no or low concentration of producing effect. 

## Explore your data (Litter)

12. Determine the class of collectDate. Is it a date? If not, change to a date and confirm the new class of the variable. Using the `unique` function, determine which dates litter was sampled in August 2018.

```{r}
class(Litter$collectDate)
view(Litter)
```
```{r}
Litter$collectDate <- as.Date(Litter$collectDate, format = "%Y-%m-%d")
class(Litter$collectDate)
```
```{r}
unique(Litter$collectDate)
```


13. Using the `unique` function, determine how many plots were sampled at Niwot Ridge. How is the information obtained from `unique` different from that obtained from `summary`?

```{r}
unique(Litter)
```

> Answer: There are 52 unique plots obtained by function unique and look at measuredBy column. Summary will provide unfavorable long result because it provide summary without omit duplicates.

14. Create a bar graph of functionalGroup counts. This shows you what type of litter is collected at the Niwot Ridge sites. Notice that litter types are fairly equally distributed across the Niwot Ridge sites.

```{r}
ggplot(Litter) + 
  geom_bar(aes(x = functionalGroup))
```

15. Using `geom_boxplot` and `geom_violin`, create a boxplot and a violin plot of dryMass by functionalGroup.

```{r}
ggplot(Litter) + 
  geom_boxplot(aes(x = functionalGroup, y = dryMass))
```
```{r}
ggplot(Litter) + 
  geom_violin(aes(x = functionalGroup, y = dryMass))
```


Why is the boxplot a more effective visualization option than the violin plot in this case?

> Answer: Because in this case, the data is nearly identical, so violin plot will be "squeezed" together and we cannot read any effective data. Whereas, box plot is drown by defining inter quatile range, which almost gurantee we will have that "box" only by minor difference. 

What type(s) of litter tend to have the highest biomass at these sites?

> Answer:Needles tend to have the highest biomass, it has the largest range and largest inter quatile range and the longes tail. 


