---
title: "Assignment 7: GLMs (Linear Regressios, ANOVA, & t-tests)"
author: "Xuancheng Guo"
date: "Spring 2024"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Rename this file `<FirstLast>_A07_GLMs.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

```{r setup1}
# Load packages
getwd()
library(tidyverse)
library(agricolae)
library(dplyr)
library(lubridate)
library(ggplot2)
library(here)

# Load data sets
NTL.LTER <- read.csv(here("Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv"),
                     stringsAsFactors = TRUE)
NTL.LTER$sampledate <- as.Date(NTL.LTER$sampledate, format = "%m/%d/%y")
class(NTL.LTER$sampledate)
```


2. Build a ggplot theme and set it as your default theme.

```{r setup2}
# Build Theme
mytheme <- theme_classic(base_size = 14) +
  theme(
    axis.text = element_text(color = "black"), 
    legend.position = "top", 
    plot.title = element_text(hjust = 0.5, face = "bold"), 
    axis.title = element_text(face = "bold"),
    plot.margin = margin(10, 10, 10, 10)
  )

# Call out mytheme
theme_set(mytheme)

```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:
> Answer:
H0:  The mean lake temperature recorded during July does not change with depth across all lakes.
Ha: The mean lake temperature recorded during July changes with depth across all lakes.


4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.

```{r scatterplot}
#4
NTL.LTER.new <- NTL.LTER %>%
  mutate(month = month(sampledate)) %>%
  filter(month == 7) %>%
  select(lakename, year4, daynum, depth, temperature_C) %>%
  na.omit()

#5
temp.dep <- ggplot(NTL.LTER.new, aes(x = depth, y = temperature_C)) +
  geom_point () +
  geom_smooth(method = lm) +
  ylim(0, 35)
print(temp.dep)
```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: The figure suggests that as depth increase, temperature of corresponding area will decrease, but the linear prediction migth not be very accurate since the result shows a curved drop; also, the linear prediciton cannot capture result exceed 11 (roughly).


7. Perform a linear regression to test the relationship and display the results.

```{r linear.regression}
#7
temp.dep.reg <- lm(
  data = NTL.LTER.new,
  temperature_C ~ depth
  )
summary(temp.dep.reg)

```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer: 
Looking at R-square/adjusted R-square value, 73.87% of the observation is explained by the linear model presented previously. 
The degrees of freedom is n-2 which is summarized by the code, 9726. 
The result is statistically significant in such we will reject the null hypothesis at a P-value approaching 0 and conclude "the mean lake temperature recorded during July changes with depth across all lakes." 
By looking at point estimates of depth, we have a statistically significant result, as depth increase by 1 unit below surface, temperature will be decrease by 1.946 units. 

---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9
temp.dep.AIC <- lm(
  data = NTL.LTER.new,
  temperature_C ~ year4 + daynum + depth
)

step(temp.dep.AIC)

#10
summary(temp.dep.AIC)

```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: We should use all three input as variable for this regression since all three coefficien has a statistically significant level at 1% confidence level. 
This result is an improvement than only use depth as the explanatory variable. 



---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12
temp.anova <- aov(data = NTL.LTER,
  temperature_C ~ lakename
  )
summary(temp.anova)

```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: There is significant difference in temperature among lakes due to the small test P-value, so we reject the null hypothesis and conclude there is at least 1 difference among lakes. 



14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2}
#14.
q14.graph <- ggplot(NTL.LTER.new,
  aes(x = depth, y = temperature_C, color = lakename)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE) +
  ylim(0, 35) +
  mytheme
print(q14.graph)

```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15
temp.HSD <- HSD.test(temp.anova, "lakename", group = T)
print(temp.HSD)
TukeyHSD(temp.anova)
```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer: Paul Lake and Ward Lake is likely to have the same temperature; Tuesday Lake and Hummingbird Lake also don't have a significant difference; Hummingbird Lake is also not statistically different from East Long Lake, but East Long Lake is different from Tuesday Lake. 
We could tell this simply from the report or look at all cross-comparing P-values, whichever P-value is large, means we cannot reject the null hypothesis that they have difference. 
Except the three pairs mentioned above, all other lakes are distinct than each other. 

 

17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: We can run two-sample T-test or ANOVA to test the mean difference. 



18. Wrangle the July data to include only records for Crampton Lake and Ward Lake. Run the two-sample T-test on these data to determine whether their July temperature are same or different. What does the test say? Are the mean temperatures for the lakes equal? Does that match you answer for part 16?

```{r t.test}
# Filter out our data
Cramp.Ward <- NTL.LTER.new %>%
  filter(lakename == "Crampton Lake" | lakename == "Ward Lake")
summary(Cramp.Ward$lakename)

# Two sample T
ttest <- t.test(Cramp.Ward$temperature_C ~ Cramp.Ward$lakename)
print(ttest)
```

>Answer: Because the P-value is large than 95% confidence level, we fail to reject that the two lakes have significant difference. This result is contradictory from Q16. One guess could be in Q16, we cross-compared all lakes, so this method first define what is "difference", but Q18 we only performed a two sample T which we will compare our result with simple statistics. 
