---
title: "Assignment 2: Coding Basics"
author: Xuancheng Guo
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics.

## Directions

1.  Rename this file `<FirstLast>_A02_CodingBasics.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction.
4.  Be sure to **answer the questions** in this assignment document.
5.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
6.  After Knitting, submit the completed exercise (PDF file) to Sakai.

## Basics, Part 1

1.  Generate a sequence of numbers from one to 30, increasing by threes. Assign this sequence a name.
```{r}
seq1_30_3 <- seq(1,30,3)
```
2.  Compute the mean and median of this sequence.
```{r}
mean(seq1_30_3)
median(seq1_30_3)
mean_seq1 <- mean(seq1_30_3)
median_seq1<- median(seq1_30_3)
```

3.  Ask R to determine whether the mean is greater than the median.
```{r}
mean_seq1 > median_seq1
```


4.  Insert comments in your code to describe what you are doing.

```{r basics 1}
#1. Generate a sequence and give it a name

#2. Let R find the mean and median of sequence 1 and give them a name, correspondingly

#3. This argument will return TRUE if the mean is greater than the median, but in this case, the argument is false. 

```

## Basics, Part 2

5.  Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

```{r}
# Student names
student_names <- c("Amy", "Bob", "Cat", "Dog")

# Test scores out of 100
test_scores <- c(75, 90, 48, 60)

# Passed or not
pass_status <- test_scores >= 50

# Data Frame
student_data <- data.frame(Student_Name = student_names, Test_Score = test_scores, Pass_Status = pass_status)

# Return the frame
print(student_data)
```

6.  Label each vector with a comment on what type of vector it is.

7.  Combine each of the vectors into a data frame. Assign the data frame an informative name.

8.  Label the columns of your data frame with informative titles.
#Function under 5 will do

9.  QUESTION: How is this data frame different from a matrix?

> Answer: The matrix must contain the same data type (all numbers or all text, etc). Matrix also don't need a column/row name, a "matrix name" might be okay. 

10. Create a function with an if/else statement. Your function should take a **vector** of test scores and print (not return) whether a given test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement. 
```{r}
ifelse(test_scores >= 50, "Pass", "Did not pass")
```

11. Apply your function to the vector with test scores that you created in number 5.
#Function under 10 will do

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer:In this case, only ifelse will work because we want to compare a vector with a scalar. If/else can only work with scalar. 
